DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-12 2015CompleteDerivation/completeDerivation.tex
First pass at discussion and conclusion

%Time arrow

%Particle excitations

%Mixed state as fundamental (law of evolution defined on the mixed state)

%Black hole as irreducible object

@@

>>>>

As we went through many details in the derivation, it is time to step back and conclude with a few general remark.

The first remarkable aspect of this work is that it can exists. Regardless of whether all the minor details are addressed in the best possible way, the idea that so much can be derived from so little is unexpected.

It often happened that the inability to justify one detail forced a partial shift in thinking that enabled to connect more ideas more simply. The ability to constrain the Hamiltonian to vector potential came as a surprise after we realized we needed distributions as a core concept to justify Hamilton's equations. The fact that relationships must be continuous functions came as a surprise after we connected topology with the notion of physical distinguishability. It was this constant simplification that shows us that we were on the right track and that, in the end, decided the boundary of the work as the initial goals were very modest in comparison.

The second striking aspect is a shift in perspective that this work forces upon us. The usual reductionist view is that you start with the rules of the small parts and then build up the composite whole. This works shift that perspective upside down: it's not what the parts can tell you about the whole, it's what the whole can tell you about the parts.

The state space for classical particles is the one that allows to express invariant distributions: that is a requirement that comes from the whole. The fragments of quantum particles do not even have a well defined description: the whole tells us nothing about them. In retrospect, this is not be surprising. Nobody is going to hand us down the laws of infinitesimal parts: we induce them by manipulating finite systems.

The third notable aspect is that the assumptions have more to do with our ability to create a reliable description rather than essential features of the system itself. A system is not deterministic per se: it is a particular choice for its description (i.e. state) that may be deterministic under some evolution. In the same vein, a material is not homogeneous per se. Or infinitesimally reducible. In this context, the state, its properties and the law of evolution are the description of the system under our idealization, not some intrinsic feature of the system.

The fourth intriguing aspect is that, when framing classical and quantum mechanics in this way, there is a progression. First we assume a homogeneous infinitesimally reducible material. Then we assume an homogeneous irreducible material. Connecting the dots, the next assumption would be a inhomogeneous irreducible material: where the material is not in constant equilibrium and its particles are not composed of the same amount. This different starting point may lead to the resolution of some of the problems we mentioned.

As different particle can transform into one another, they could be seen as different stable states of the same inhomogeneous material. In particular, since $\sigma_x \sigma_v \geq \frac{\hbar}{2m}$, particles in different generations (i.e. electron vs muon) could be similar stable configurations where the fragments of the material are more "packed" in terms of states labeled by the kinematic variables.

As we are not always at equilibrium, the state of the system will be in general a statistical mixture of distributions of material. But the evolution may also take pure states (i.e. a mixed state with a single state with $100\%$ probability) and evolve them into mixed states, thus giving an account of what happens during a measurement or a decay.

The law of evolution may still be deterministic and reversible, in the sense that given a particular mixed state we can predict/reconstruct future/past mixed state, but it will not be time symmetric.  We can then tell that time flows in the direction where information is exchanged, where entropy increases. This would give us a time arrow.

These ideas are appealing because they seem to fit well in the narrative, but at this point they are just speculation. The only thing we can say at this point is that the state space would be a vector space on a field that is neither $\mathbbm{R}$ or $\mathbbm{C}$.

In the near future it would seem appropriate to extend this work to the remaining part of well-established physics. The ideas toward that goal are more concrete. The state space for a quantized material, already introduced here, should be captured by the Fock space. The idea of a non-minimal material will force us to consider vector and tensor quantities at each point in the distribution, whose state should be captured by spin systems. Assuming that the vector potential forces themselves are dependent on the state of a material should hopefully give us the Hamiltonian for the fields.

@@

We have seen that much of fundamental physics can be derived from a few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of material. The main physical concepts can be summarized by the following points:

\begin{itemize}

\item The state space is not just a collection of states but also captures its physical properties such as physical distinguishability (represented by a topology), the ability to count states (represented by a measure, symplectic form or metric tensor) or decomposability (represented by a vector space structure).

\item Deterministic and reversible evolution is a one-to-one map that preserves the nature of the system (e.g. physical distinguishability, state count and decomposability). Therefore it is an isomorphism on the mathematical structure defined by the state space (e.g. homeomorphism, canonical transformation, unitary transformation), not just a state-to-state bijection.

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material for which the state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is the symplectic manifold $(T^*\mathcal{Q}, \omega)$; deterministic and reversible evolution is a symplectomorphism (i.e. a canonical transformation).

\item If the classical particle states are fully characterized by their motion in space, then the states also evolve according to Lagrangian mechanics and the motion is that of a massive particle under scalar/vector potential forces.

\item Relativistic motion arises by properly handling time dependent deterministic and reversible evolution for the material (i.e. distributions over phase space).

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is an inner vector product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, the feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and it often gave intuitive ideas that connects to different branches of math and physics: it cannot be a simple coincidence.

While the scope of this work is limited to re-founding and re-understanding the already known physical theories, it is our hope that this insight may also lead to new ideas to address some of the open problems in the foundations of physics.
